---
title: "mr460-week7"
format: html
execute:
  echo: false
  warning: false
  message: false
---

# WEEK 7 - Feedforward Neural Networks

## Logistic Regression vs MLP

## ReLU vs tanh

## Capacity + Regularisation

```{r}
set.seed(460)

if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")

library(ggplot2)
library(readr)
```

```{r}
file_candidates <- c(
  "heart_failure_clinical_records_dataset.csv",
  "data/heart_failure_clinical_records_dataset.csv"
)

file_path <- file_candidates[file.exists(file_candidates)][1]

if (!is.na(file_path)) {
  df <- read_csv(file_path, show_col_types = FALSE)
} else {
  url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00519/heart_failure_clinical_records_dataset.csv"
  df <- tryCatch(
    read_csv(url, show_col_types = FALSE),
    error = function(e) {
      stop("Could not load the dataset. Save the CSV into your project folder and try again.")
    }
  )
}

df$DEATH_EVENT <- as.numeric(df$DEATH_EVENT)

cat("Rows:", nrow(df), " Columns:", ncol(df), "\n")
cat("Class balance:\n")
print(table(df$DEATH_EVENT))
```

```{r}
stratified_split <- function(data, target = "DEATH_EVENT", train_prop = 0.70, val_prop_within_train = 0.80) {
  idx0 <- which(data[[target]] == 0)
  idx1 <- which(data[[target]] == 1)

  train0 <- sample(idx0, size = floor(length(idx0) * train_prop))
  train1 <- sample(idx1, size = floor(length(idx1) * train_prop))

  test_idx <- setdiff(seq_len(nrow(data)), c(train0, train1))
  train_full_idx <- c(train0, train1)

  train_full <- data[train_full_idx, ]
  test_df <- data[test_idx, ]

  idx0_tf <- which(train_full[[target]] == 0)
  idx1_tf <- which(train_full[[target]] == 1)

  train_inner0 <- sample(idx0_tf, size = floor(length(idx0_tf) * val_prop_within_train))
  train_inner1 <- sample(idx1_tf, size = floor(length(idx1_tf) * val_prop_within_train))

  train_idx <- c(train_inner0, train_inner1)
  val_idx <- setdiff(seq_len(nrow(train_full)), train_idx)

  list(
    train = train_full[train_idx, ],
    val   = train_full[val_idx, ],
    test  = test_df
  )
}

splits <- stratified_split(df)

train_df <- splits$train
val_df   <- splits$val
test_df  <- splits$test

cat("\nTrain class balance:\n")
print(prop.table(table(train_df$DEATH_EVENT)))
cat("\nValidation class balance:\n")
print(prop.table(table(val_df$DEATH_EVENT)))
cat("\nTest class balance:\n")
print(prop.table(table(test_df$DEATH_EVENT)))
```

```{r}
feature_names <- setdiff(names(df), "DEATH_EVENT")

x_train_raw <- as.matrix(train_df[, feature_names])
x_val_raw   <- as.matrix(val_df[, feature_names])
x_test_raw  <- as.matrix(test_df[, feature_names])

y_train <- matrix(train_df$DEATH_EVENT, ncol = 1)
y_val   <- matrix(val_df$DEATH_EVENT, ncol = 1)
y_test  <- matrix(test_df$DEATH_EVENT, ncol = 1)

train_means <- colMeans(x_train_raw)
train_sds <- apply(x_train_raw, 2, sd)
train_sds[train_sds == 0] <- 1

scale_with_train <- function(x, means, sds) {
  scale(x, center = means, scale = sds)
}

x_train <- scale_with_train(x_train_raw, train_means, train_sds)
x_val   <- scale_with_train(x_val_raw, train_means, train_sds)
x_test  <- scale_with_train(x_test_raw, train_means, train_sds)
```

```{r}
sigmoid <- function(z) 1 / (1 + exp(-pmin(pmax(z, -30), 30)))

binary_accuracy <- function(y_true, y_prob, threshold = 0.5) {
  y_pred <- ifelse(y_prob >= threshold, 1, 0)
  mean(y_pred == y_true)
}

roc_auc_manual <- function(y_true, y_prob) {
  y_true <- as.numeric(y_true)
  y_prob <- as.numeric(y_prob)

  pos <- y_prob[y_true == 1]
  neg <- y_prob[y_true == 0]

  if (length(pos) == 0 || length(neg) == 0) return(NA_real_)

  ranks <- rank(c(pos, neg))
  pos_ranks <- ranks[seq_along(pos)]
  (sum(pos_ranks) - length(pos) * (length(pos) + 1) / 2) / (length(pos) * length(neg))
}

bce_loss <- function(y_true, y_prob) {
  eps <- 1e-8
  y_prob <- pmin(pmax(y_prob, eps), 1 - eps)
  -mean(y_true * log(y_prob) + (1 - y_true) * log(1 - y_prob))
}
```

# Task 1

```{r}
lr_fit <- glm(DEATH_EVENT ~ ., data = train_df, family = binomial())

val_prob_lr  <- predict(lr_fit, newdata = val_df, type = "response")
test_prob_lr <- predict(lr_fit, newdata = test_df, type = "response")

lr_results <- data.frame(
  Model = "Logistic Regression",
  Split = c("Validation", "Test"),
  Accuracy = c(binary_accuracy(y_val, val_prob_lr), binary_accuracy(y_test, test_prob_lr)),
  ROC_AUC = c(roc_auc_manual(y_val, val_prob_lr), roc_auc_manual(y_test, test_prob_lr))
)

lr_results
```

```{r}
relu <- function(z) {
  z <- as.matrix(z)
  z[z < 0] <- 0
  z
}

relu_deriv <- function(z) {
  z <- as.matrix(z)
  out <- matrix(0, nrow = nrow(z), ncol = ncol(z))
  out[z > 0] <- 1
  out
}

tanh_act <- function(z) {
  z <- as.matrix(z)
  tanh(z)
}

tanh_deriv <- function(z) {
  z <- as.matrix(z)
  1 - tanh(z)^2
}

init_model <- function(input_dim, hidden_units) {
  list(
    W1 = matrix(rnorm(input_dim * hidden_units, sd = 0.1),
                nrow = input_dim, ncol = hidden_units),
    b1 = rep(0, hidden_units),
    W2 = matrix(rnorm(hidden_units, sd = 0.1),
                nrow = hidden_units, ncol = 1),
    b2 = 0
  )
}

predict_mlp <- function(model, X, activation = "relu") {
  X <- as.matrix(X)
  W1 <- as.matrix(model$W1)
  W2 <- as.matrix(model$W2)

  Z1 <- sweep(X %*% W1, 2, model$b1, FUN = "+")
  if (activation == "relu") {
    A1 <- relu(Z1)
  } else {
    A1 <- tanh_act(Z1)
  }

  A1 <- matrix(A1, nrow = nrow(X))
  Z2 <- A1 %*% W2 + model$b2
  as.numeric(sigmoid(Z2))
}

train_mlp <- function(
  X_train, y_train, X_val, y_val,
  hidden_units = 32,
  activation = "relu",
  epochs = 300,
  lr = 0.01,
  l2 = 0,
  seed = 460
) {
  set.seed(seed)

  X_train <- as.matrix(X_train)
  X_val <- as.matrix(X_val)
  y_train <- matrix(as.numeric(y_train), ncol = 1)
  y_val <- matrix(as.numeric(y_val), ncol = 1)

  n <- nrow(X_train)
  p <- ncol(X_train)

  model <- init_model(p, hidden_units)

  history <- data.frame(
    epoch = integer(),
    train_loss = numeric(),
    val_loss = numeric(),
    train_acc = numeric(),
    val_acc = numeric()
  )

  for (epoch in 1:epochs) {
    W1 <- as.matrix(model$W1)
    W2 <- as.matrix(model$W2)

    Z1 <- sweep(X_train %*% W1, 2, model$b1, FUN = "+")

    if (activation == "relu") {
      A1 <- relu(Z1)
      dA1_dZ1 <- relu_deriv(Z1)
    } else {
      A1 <- tanh_act(Z1)
      dA1_dZ1 <- tanh_deriv(Z1)
    }

    A1 <- matrix(A1, nrow = n)
    dA1_dZ1 <- matrix(dA1_dZ1, nrow = n)

    Z2 <- A1 %*% W2 + model$b2
    A2 <- matrix(sigmoid(Z2), ncol = 1)

    dZ2 <- A2 - y_train
    dW2 <- (t(A1) %*% dZ2) / n + l2 * W2
    db2 <- mean(dZ2)

    dA1 <- dZ2 %*% t(W2)
    dZ1 <- dA1 * dA1_dZ1

    dW1 <- (t(X_train) %*% dZ1) / n + l2 * W1
    db1 <- colMeans(dZ1)

    model$W1 <- W1 - lr * dW1
    model$b1 <- model$b1 - lr * db1
    model$W2 <- W2 - lr * dW2
    model$b2 <- model$b2 - lr * db2

    train_prob <- predict_mlp(model, X_train, activation)
    val_prob   <- predict_mlp(model, X_val, activation)

    history <- rbind(
      history,
      data.frame(
        epoch = epoch,
        train_loss = bce_loss(y_train, train_prob),
        val_loss = bce_loss(y_val, val_prob),
        train_acc = binary_accuracy(y_train, train_prob),
        val_acc = binary_accuracy(y_val, val_prob)
      )
    )
  }

  list(model = model, history = history)
}

plot_learning_curves <- function(history, title_text) {
  p1 <- ggplot(history, aes(x = epoch)) +
    geom_line(aes(y = train_loss, colour = "Train Loss")) +
    geom_line(aes(y = val_loss, colour = "Validation Loss")) +
    labs(title = paste(title_text, "- Loss"), y = "Loss", colour = "") +
    theme_minimal()

  p2 <- ggplot(history, aes(x = epoch)) +
    geom_line(aes(y = train_acc, colour = "Train Accuracy")) +
    geom_line(aes(y = val_acc, colour = "Validation Accuracy")) +
    labs(title = paste(title_text, "- Accuracy"), y = "Accuracy", colour = "") +
    theme_minimal()

  print(p1)
  print(p2)
}

summarise_model <- function(model_name, fit_obj, X_val, y_val, X_test, y_test, activation) {
  val_prob <- predict_mlp(fit_obj$model, X_val, activation)
  test_prob <- predict_mlp(fit_obj$model, X_test, activation)

  data.frame(
    Model = model_name,
    Split = c("Validation", "Test"),
    Accuracy = c(binary_accuracy(y_val, val_prob), binary_accuracy(y_test, test_prob)),
    ROC_AUC = c(roc_auc_manual(y_val, val_prob), roc_auc_manual(y_test, test_prob))
  )
}
```

```{r}
mlp_relu_fit <- train_mlp(
  X_train = x_train,
  y_train = y_train,
  X_val = x_val,
  y_val = y_val,
  hidden_units = 32,
  activation = "relu",
  epochs = 300,
  lr = 0.01
)

summarise_model("MLP ReLU (32)", mlp_relu_fit, x_val, y_val, x_test, y_test, "relu")
plot_learning_curves(mlp_relu_fit$history, "MLP ReLU (32 units)")
```

# Task 2

```{r}
mlp_tanh_fit <- train_mlp(
  X_train = x_train,
  y_train = y_train,
  X_val = x_val,
  y_val = y_val,
  hidden_units = 32,
  activation = "tanh",
  epochs = 300,
  lr = 0.01
)

rbind(
  summarise_model("MLP ReLU (32)", mlp_relu_fit, x_val, y_val, x_test, y_test, "relu"),
  summarise_model("MLP tanh (32)", mlp_tanh_fit, x_val, y_val, x_test, y_test, "tanh")
)

plot_learning_curves(mlp_tanh_fit$history, "MLP tanh (32 units)")
```

# Task 3

```{r}
mlp_small_fit <- train_mlp(
  X_train = x_train,
  y_train = y_train,
  X_val = x_val,
  y_val = y_val,
  hidden_units = 8,
  activation = "relu",
  epochs = 300,
  lr = 0.01
)

mlp_large_fit <- train_mlp(
  X_train = x_train,
  y_train = y_train,
  X_val = x_val,
  y_val = y_val,
  hidden_units = 128,
  activation = "relu",
  epochs = 300,
  lr = 0.01
)

mlp_large_l2_fit <- train_mlp(
  X_train = x_train,
  y_train = y_train,
  X_val = x_val,
  y_val = y_val,
  hidden_units = 128,
  activation = "relu",
  epochs = 300,
  lr = 0.01,
  l2 = 0.001
)

rbind(
  summarise_model("Small MLP (8)", mlp_small_fit, x_val, y_val, x_test, y_test, "relu"),
  summarise_model("Large MLP (128)", mlp_large_fit, x_val, y_val, x_test, y_test, "relu"),
  summarise_model("Large MLP + L2 (128)", mlp_large_l2_fit, x_val, y_val, x_test, y_test, "relu")
)

plot_learning_curves(mlp_small_fit$history, "Small MLP (8 units)")
plot_learning_curves(mlp_large_fit$history, "Large MLP (128 units)")
plot_learning_curves(mlp_large_l2_fit$history, "Large MLP + L2 (128 units)")
```

# Task 4

```{r}
final_candidates <- data.frame(
  Model = c(
    "Logistic Regression",
    "MLP ReLU (32)",
    "MLP tanh (32)",
    "Small MLP (8)",
    "Large MLP (128)",
    "Large MLP + L2 (128)"
  ),
  Validation_AUC = c(
    lr_results$ROC_AUC[lr_results$Split == "Validation"],
    summarise_model("MLP ReLU (32)", mlp_relu_fit, x_val, y_val, x_test, y_test, "relu")$ROC_AUC[1],
    summarise_model("MLP tanh (32)", mlp_tanh_fit, x_val, y_val, x_test, y_test, "tanh")$ROC_AUC[1],
    summarise_model("Small MLP (8)", mlp_small_fit, x_val, y_val, x_test, y_test, "relu")$ROC_AUC[1],
    summarise_model("Large MLP (128)", mlp_large_fit, x_val, y_val, x_test, y_test, "relu")$ROC_AUC[1],
    summarise_model("Large MLP + L2 (128)", mlp_large_l2_fit, x_val, y_val, x_test, y_test, "relu")$ROC_AUC[1]
  )
)

final_candidates
```

```{r}
best_model_name <- final_candidates$Model[which.max(final_candidates$Validation_AUC)]
best_model_name
```

```{r}
if (best_model_name == "Logistic Regression") {
  final_test_acc <- lr_results$Accuracy[lr_results$Split == "Test"]
  final_test_auc <- lr_results$ROC_AUC[lr_results$Split == "Test"]
} else if (best_model_name == "MLP ReLU (32)") {
  tmp <- summarise_model("MLP ReLU (32)", mlp_relu_fit, x_val, y_val, x_test, y_test, "relu")
  final_test_acc <- tmp$Accuracy[2]
  final_test_auc <- tmp$ROC_AUC[2]
} else if (best_model_name == "MLP tanh (32)") {
  tmp <- summarise_model("MLP tanh (32)", mlp_tanh_fit, x_val, y_val, x_test, y_test, "tanh")
  final_test_acc <- tmp$Accuracy[2]
  final_test_auc <- tmp$ROC_AUC[2]
} else if (best_model_name == "Small MLP (8)") {
  tmp <- summarise_model("Small MLP (8)", mlp_small_fit, x_val, y_val, x_test, y_test, "relu")
  final_test_acc <- tmp$Accuracy[2]
  final_test_auc <- tmp$ROC_AUC[2]
} else if (best_model_name == "Large MLP (128)") {
  tmp <- summarise_model("Large MLP (128)", mlp_large_fit, x_val, y_val, x_test, y_test, "relu")
  final_test_acc <- tmp$Accuracy[2]
  final_test_auc <- tmp$ROC_AUC[2]
} else {
  tmp <- summarise_model("Large MLP + L2 (128)", mlp_large_l2_fit, x_val, y_val, x_test, y_test, "relu")
  final_test_acc <- tmp$Accuracy[2]
  final_test_auc <- tmp$ROC_AUC[2]
}

data.frame(
  Selected_Model = best_model_name,
  Test_Accuracy = final_test_acc,
  Test_ROC_AUC = final_test_auc
)
```

# Short analysis notes

- Logistic regression is linear because it models the log-odds as a linear combination of the input features.
- If the MLP does better on validation metrics, that suggests the hidden layer captures nonlinear relationships.
- ReLU often converges faster, while tanh can be slower because gradients shrink near saturation.
- Larger networks have more capacity and can overfit more easily on small datasets.
- L2 regularisation helps reduce overfitting by discouraging very large weights.
- Training data fits the model, validation data selects the model, and the test set is used once at the end for final evaluation.