---
title: "mr460-week5-clustering"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    df-print: paged
  pdf:
    toc: true
    number-sections: true
editor: source
---
```{r}
library(tidyverse)
library(lubridate)
library(here)

# clustering helpers
library(cluster)   # silhouette()
set.seed(42)
```

## Step 1 — Unit of analysis

**Unit of analysis:** **customers** (one row per customer).

**Why this unit makes sense for segmentation:**  
Customer segmentation aims to group people with similar purchasing behaviour so we can describe “types” of shoppers (e.g., frequent small-basket buyers vs. rare high spenders). The Online Retail data is transactional (many rows per customer), so clustering requires us to **aggregate** transactions into a compact customer “behaviour profile” (features) before applying distance-based clustering.

**Key limitation:** customers with missing `Customer ID` cannot be clustered (they have no stable identifier), and aggregation can hide within-customer variability (e.g., a customer who changes behaviour over time).

##Step 2 — Feature construction (customer behaviour profiles)

We keep a **purchases-only** dataset by removing cancellations (Invoices starting with `"C"`), and non-sensical values (`Quantity <= 0` or `Price <= 0`).

```{r load-clean, message=FALSE, warning=FALSE}
df_raw <- readr::read_csv(
  here("data", "raw", "online_retail_II(4).csv"),
  show_col_types = FALSE
)

df1 <- df_raw %>%
  mutate(
    InvoiceDate = dmy_hms(InvoiceDate),
    TotalPrice  = Quantity * Price,
    is_cancel   = str_starts(Invoice, "C")
  )

df_purchases <- df1 %>%
  filter(!is_cancel, Quantity > 0, Price > 0) %>%
  filter(!is.na(InvoiceDate))

# Keep only customers with IDs (needed for customer-level aggregation)
df_purchases_id <- df_purchases %>% filter(!is.na(`Customer ID`))

glimpse(df_purchases_id)
```
We create **one row per customer** and compute behaviour features:

- **Recency (days):** how many days since last purchase (relative to dataset end).
- **Frequency:** number of invoices (repeat buying).
- **Monetary value:** total historical spend.
- Additional structure: basket size/value, product variety.

```{r make-customer-features, message=FALSE, warning=FALSE}
cutoff_date <- max(df_purchases_id$InvoiceDate, na.rm = TRUE)

cust <- df_purchases_id %>%
  group_by(`Customer ID`) %>%
  summarise(
    n_invoices       = n_distinct(Invoice),              # frequency proxy
    n_items          = n(),                              # engagement proxy
    n_products       = n_distinct(StockCode),            # variety proxy
    total_spend      = sum(TotalPrice, na.rm = TRUE),    # monetary proxy
    avg_item_price   = mean(Price, na.rm = TRUE),
    avg_basket_value = total_spend / n_invoices,
    avg_items_per_invoice = n_items / n_invoices,
    last_purchase    = max(InvoiceDate, na.rm = TRUE),
    recency_days     = as.numeric(difftime(cutoff_date, last_purchase, units = "days")),
    active_span_days = as.numeric(difftime(max(InvoiceDate), min(InvoiceDate), units = "days")),
    .groups = "drop"
  ) %>%
  select(-last_purchase)

summary(cust)
```
### Skew handling (practical)

Spend-like features are often extremely right-skewed. For clustering, this can cause a few large values to dominate distances even after scaling. A common, transparent fix is `log1p()` (log(1+x)) for strictly-positive, heavy-tailed features.

```{r skew-fix, message=FALSE, warning=FALSE}
cust_feat <- cust %>%
  mutate(
    total_spend_log = log1p(total_spend),
    avg_basket_value_log = log1p(avg_basket_value),
    n_invoices_log = log1p(n_invoices),
    n_products_log = log1p(n_products),
    n_items_log = log1p(n_items)
  )

cust_feat %>%
  select(`Customer ID`, total_spend, total_spend_log, n_invoices, n_invoices_log) %>%
  slice_head(n = 5)
```
##Step 3 — Scaling (why it matters)

Clustering here is **distance-based**. If features are on different scales (e.g., `total_spend` in thousands while `recency_days` is in tens), then **large-scale features dominate** Euclidean distance and determine clusters regardless of the others.

**Method used:** standardisation (z-scores) with `scale()`.

**What would happen without scaling:** clusters would mostly separate customers by the largest-range features (typically spend), and recency/frequency would contribute little.

```{r scale-features, message=FALSE, warning=FALSE}
# Choose a compact feature set for clustering
X <- cust_feat %>%
  transmute(
    recency_days,
    n_invoices_log,
    total_spend_log,
    avg_basket_value_log,
    n_products_log
  )

# Standardise (mean 0, sd 1)
X_scaled <- scale(X)

# sanity check: means ~0, sds ~1
apply(X_scaled, 2, mean) %>% round(3)
apply(X_scaled, 2, sd) %>% round(3)
```
##Step 4 — Apply clustering

### 4A — K-means (required)
#### Choosing *k* (avoid arbitrary decisions)

K-means requires a number of clusters *k*. There is no single “correct” value, but we can make an informed choice using:

- **Elbow / within-cluster sum of squares (WSS):** diminishing returns as k increases.
- **Silhouette:** how well-separated clusters are (higher is better), but we treat it as guidance, not a goal.

```{r choose-k, message=FALSE, warning=FALSE}
wss <- map_dbl(2:10, ~kmeans(X_scaled, centers = .x, nstart = 25)$tot.withinss)

sil <- map_dbl(2:10, function(k){
  km <- kmeans(X_scaled, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(X_scaled))
  mean(ss[, 3])
})

k_diagnostics <- tibble(k = 2:10, wss = wss, silhouette = sil)
k_diagnostics
```

```{r plot-k, message=FALSE, warning=FALSE}
ggplot(k_diagnostics, aes(k, wss)) +
  geom_line() + geom_point() +
  labs(title = "Elbow plot (WSS)", x = "k", y = "Total within-cluster SS")

ggplot(k_diagnostics, aes(k, silhouette)) +
  geom_line() + geom_point() +
  labs(title = "Average silhouette by k", x = "k", y = "Average silhouette")
```

**Chosen k:** I use **k = 4** as a compromise between (1) clear diminishing returns in WSS after a small number of clusters and (2) a reasonable silhouette without producing tiny clusters that are hard to interpret.  
(If your plots clearly suggest a different k, you can change it here and explain why.)

```{r kmeans-fit, message=FALSE, warning=FALSE}
k <- 4
km <- kmeans(X_scaled, centers = k, nstart = 50)

cust_km <- cust_feat %>%
  mutate(cluster = factor(km$cluster))

cust_km %>% count(cluster) %>% arrange(desc(n))
```

### 4B — Inspect cluster centres (interpretation-friendly)

K-means centres are in **scaled units**, so to interpret we compare **original-scale summaries** by cluster.

```{r cluster-profiles, message=FALSE, warning=FALSE}
profiles <- cust_km %>%
  group_by(cluster) %>%
  summarise(
    n_customers = n(),
    recency_days_mean = mean(recency_days),
    n_invoices_mean = mean(n_invoices),
    total_spend_mean = mean(total_spend),
    avg_basket_value_mean = mean(avg_basket_value),
    n_products_mean = mean(n_products),
    .groups = "drop"
  ) %>%
  arrange(desc(n_customers))

profiles
```

### 4C — Visualise clusters with PCA (2D view)

We reduce the (scaled) feature space to 2 principal components for a **visual diagnostic** (not proof of correctness).

```{r pca-plot, message=FALSE, warning=FALSE}
pca <- prcomp(X_scaled, center = FALSE, scale. = FALSE)

pca_df <- as_tibble(pca$x[, 1:2]) %>%
  setNames(c("PC1", "PC2")) %>%
  mutate(cluster = cust_km$cluster)

ggplot(pca_df, aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.6) +
  labs(title = "Customer clusters (K-means) shown in PCA space")
```
##Hierarchical clustering (comparison)

Hierarchical clustering can reveal whether the partitioning looks similar under a different assumption (agglomerative merging).

```{r hierarchical, message=FALSE, warning=FALSE}
d <- dist(X_scaled)
hc <- hclust(d, method = "ward.D2")

# Cut to k clusters for a direct comparison
hc_clusters <- cutree(hc, k = k)

cust_hc <- cust_feat %>%
  mutate(hc_cluster = factor(hc_clusters))

cust_hc %>% count(hc_cluster) %>% arrange(desc(n))
```
##Step 5 — Evaluate and interpret (critical discussion)

Use the profile table above to describe clusters in plain language. A typical pattern is:

- **Low recency (recent buyers) + higher frequency/value** → “active core customers”
- **High recency (haven’t bought recently) + low frequency/value** → “lapsed / one-off customers”
- **High basket value but low frequency** → “occasional big spenders”
- **High variety** → “browsers / diverse purchasers”

**Write your final interpretation using your actual `profiles` output** (cluster numbers are arbitrary; interpret by their statistics).

### Are clusters balanced?

Discuss whether one cluster dominates (often there is a very large “low-activity” cluster). If extremely imbalanced, note that k-means can favour spherical clusters and may compress small but meaningful segments.

### Sensitivity + limitations

- **Parameter sensitivity:** K-means depends on *k* and initialisation. We reduced this with `nstart = 50`, but different *k* can change the story.
- **Feature sensitivity:** clustering results are driven by feature choices. Adding/removing recency or log transforms can change geometry.
- **Scaling dependence:** without scaling, `total_spend` would dominate and clusters would mostly reflect spend alone.
- **Business meaning:** clusters may reflect data-collection artefacts (e.g., observation window, missing IDs) rather than real “types”.
